{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-12-13 sftrends_manu.Rnw
more cites to intro

Environmental condition is affected my multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}. Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.  Visual examination of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables. Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years. Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes. Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and on charactericstics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Hirsch14,Hirsch15}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP} and tidally-infleucned time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}. These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.

The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere.  Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.  Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed historical trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.  The unique ecological and social context of the Bay, including a rich source of monitoring data from the last four decades, provides a valuable opportunity to gain insight into ecosystem properties of estuaries. The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay. A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades.  Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport suggesting an approach that explicitly considers flow effects is critical for trend analysis.  To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
%\citep{Jassby00,Jassby02,Jassby08}

>>>>

Environmental condition is affected my multiple processes acting in a hierarchy across space and time \citep{Oneill89,Levin92}. Trend analysis is a broad discipline that has been applied to evaluate environmental condition using time series for the interpretation of ecologically-relevant changes.  Direct evaluation of an observed time series is often insufficient given that a long-term change can be masked by variation at shorter time scales or the observed variation represents the combined effects of many variables. Climate, local, regional, and historical effects may act individually or together to impose a change on time series, such that methods that account for variation at different scales have been used for trend analysis \citep[\cref{fig:schematic},][]{Bhangu97,Champely97,Chang08,Halliday12}.  As a practical approach for water quality evaluation, trend analysis of eutrophication endpoints often focuses on tracking the change in concentrations or loads of nutrients over many years. Indicators of eutrophication can vary naturally with variation in flow conditions and may also reflect long-term effects of management or policy changes. For example, \ac{chla} concentration as a measure of phytoplankton response to nutrient inputs can follow seasonal patterns with cyclical variation in temperature and light changes throughout each year, whereas annual trends can follow long-term variation in nutrient inputs to the system \citep{Cloern96,Cloern10}. Similarly, nutrient trends that vary with hydrologic loading also vary as a function of utilization rates by primary producers or decomposition processes \citep{Sakamoto89,Schultz08,Harding16}. Time series analysis of water quality indicators must simultaneously consider effects of processes at multiple scales and interactions between variables of interest to develop a more comprehensive description of system change.

Appropriate methods for the analysis of change depend largely on the question of interest and on charactericstics of the environmental dataset.  Trend analyses for aquatic systems have traditionally focused on comparisons between discrete periods of time to estimate a direction and magnitude of a trend using non-parametric tests \citep{Hirsch91,Esterby96}. Development of these conventional approaches addressed limitations in historical monitoring datasets related to infrequent sampling and relatively few years of continuous data. Increased availability of multi-decadal datasets, particularly for high profile environments, has prompted the development of trend analysis methods that leverage the descriptive potential of long-term time series from continuous monitoring programs \citep{Bowes09,Halliday12}. These methods are often data-driven where the parameterization of a simple functional model can change smoothly over time given that relationships between water quality variables and potential drivers are dynamic.  The \ac{wrtds} approach was developed under this context and has been used to characterize decadal trends in running-water systems \citep{Hirsch10,Sprague11,Medalie12,Hirsch14,Zhang16}. This method has the potential to provide a spatially and temporally robust description of trends by fitting a dynamic model with parameters that change relative to the domain of interest.  More recently, the \ac{wrtds} method was adapted for trend analysis in tidal waters, with a focus on \ac{chla} trends in Tampa Bay \citepalias{Beck15} and the Patuxent River Estuary \citepalias{BeckIP}, and tidally-infleunced time series of dissolved oxygen from continuous sonde measurements \citep{Beck15b}. These studies have demonstrated potential for the use \ac{wrtds} for trend analysis in tidal waters and further application to alternative datasets could provide additional insight into drivers of change in aquatic systems.

The \ac{sfe} on the Pacific Coast of the United States is one of the most prominent and culturally significant estuaries in the western hemisphere \citep{Cloern15}.  Background nutrient concentrations in the Bay often exceed those associated with excessive primary production, although eutrophication events have historically been infrequent.  Recent changes in response to additional stressors (e.g., variation in freshwater inputs/withdrawals, invasive species, climate change) suggests that Bay condition has not followed past trajectories and more subtle spatial and temporal variation could provide clues that describe underlying properties of this system \citep{Cloern12b}.  The unique ecological and social context of the Bay provides a valuable opportunity to gain insight into ecosystem properties of estuaries that define water quality dynamics at different scales. The Delta region of \ac{sfe} in particular is a mosaic of inflows that receives and processes inputs from the larger watershed to the lower Bay \citep{Jassby00,Jassby02,Jassby08}. A comprehensive monitoring dataset has been collected at several fixed locations in the Delta for the last four decades \citep{Jabusch16}.  Morever, nutrient dynamics in the Delta are inherently linked to flow variation from inputs, withdrawal, impoundments, and downstream transport \citep{Novick15}, suggesting an approach that explicitly considers flow effects is critical for trend analysis.  To date, the Delta monitoring dataset is an under-utitilized data source and a comprehensive analysis with \ac{wrtds} could facilitate an understanding of historical and recent changes in \ac{sfe} water quality.
